\section{Lecture 18 — Midterm Test}

\subsection*{Part A}

Do all of Questions 1-6.

\begin{enumerate}[label={\color{main}\sffamily\bfseries \arabic*.}]
	\item \textit{(2 pts)} Write out all the non-isomorphic abelian groups of order 36.
	\item \textit{(2 pts)} Find a composition series for the group $\mathbb Z_{36}$. Justify your answer.
	\item \textit{(2 pts)} Let $G$ be a finite group with $|G|=36$. How many Sylow 3-subgroups can $G$ have?
	\item \textit{(2 pts)} Iris is a student in MATH 4GR3 and she needs to find the class equation for a group $G$ with $|G|=36$. She shows you her answer:
	$$36=18+9+5+4.$$
	Although you don't know $G$, why do you know that Iris has an incorrect answer?
	\item \textit{(2 pts)} The following statement is false: ``If $G$ is a group with $|G|=36$, and since $18\mid 36$, then $G$ must have an element of order 18.'' Give an example to show that this statement is false.
	\item \textit{(2 pts)} Prove that there is no simple group of order $190=2\cdot 5\cdot 19$.
\end{enumerate}

\subsection*{Part B}

Do three of Questions 7-10.

\begin{enumerate}[resume,label={\color{main}\sffamily\bfseries \arabic*.}]
	\item \textit{(5 pts)} Let $G$ be any finite abelian group with $|G|=p_1^{a_1}p_2^{a_2}\cdots p_r^{a_r}$, with $a_i\geq 1$. Prove that $G$ has a subgroup of order $p_1p_2\cdots p_r$.
	\item \textit{(5 pts)} Suppose that $G$ is a finite group with subnormal series
	$$G=H_n\supset H_{n-1}\supset\cdots\supset H_1\supset H_0=\{e\},$$
	where $H_i$ is normal in $H_{i+1}$. Suppose that $|H_{i+1}/H_i|=s_{i+1}$ for $i=0,1,\hdots,n-1$. Prove that $|G|=s_1s_2\cdots s_n$.
	\item \textit{(5 pts)} Let $G$ be a finite group such that the prime $p$ divides $|G|$. Show that if $\{P_1,\hdots, P_s\}$ are all the distinct Sylow $p$-subgroups of $G$, then for any $g\in G$,
	$$\{gP_1g^{-1},\hdots,gP_sg^{-1}\}=\{P_1,\hdots, P_s\}.$$
	That is, show that $\{gP_1g^{-1},\hdots,gP_sg^{-1}\}=\{P_1,\hdots, P_s\}$ is also the set of all Sylow $p$-subgroups of $G$, but in a ``shuffled'' order.
	\item \textit{(5 pts)} Let $G$ be a finite group with $|G|=p^rm$ with $p\not\mid m$. Suppose $P_1,\hdots,P_s$ are all the distinct Sylow $p$-subgroups of $G$. Prove that $P_1\cap P_2\cap\cdots P_s$ is a normal subgroup of $G$.

	\textit{Hint.} The previous question will help. Even if you don't do Question 9, you can assume it is true for this question.
\end{enumerate}