\section{Lecture 17 — Midterm Review Questions}

\subsection*{\llap{\textcolor{main}{I.}\hspace{0.75em}}Finite Abelian Groups}

\begin{enumerate}[label=\textbf{\sffamily\color{main}\Alph*.}]
	\item Find all abelian groups of order 108.
	\item Show that there are two abelian groups of order 108 with an element of order 54.
	\item Suppose $G$ is a finite abelian group such that 10 divides $|G|$. Prove that $G$ has a cyclic subgroup of order 10.
	\item Give an example of an abelian group such that 4 divides $|G|$, but $G$ has no cyclic subgroup of order 4.
\end{enumerate}

\subsection*{\llap{\textcolor{main}{II.}\hspace{0.75em}}Solvable Groups}

\begin{enumerate}[label=\textbf{\sffamily\color{main}\Alph*.}]
	\item Find a composition series of $\mathbb Z_{20}$.
	\item Suppose $G$ has a series of subgroups
	$$G=P_n\supset P_{n-1}\supset\cdots\supset P_1\supset P_0=\{e\},$$
	where $P_i$ is normal in $P_{i+1}$ and $|P_{i+1}/P_i|$ is a prime number. Prove that $G$ is a solvable group.
\end{enumerate}

\subsection*{\llap{\textcolor{main}{III.}\hspace{0.75em}}The Characteristic Equation}

\begin{enumerate}[label=\textbf{\sffamily\color{main}\Alph*.}]
	\item Suppose $|G|=20$. Explain why
	$$20=1+2+3+6+10$$
	is not a valid class equation.
	\item Suppose $G$ is an abelian group with $|G|=2024$. What is its class equation.
\end{enumerate}

\subsection*{\llap{\textcolor{main}{IV.}\hspace{0.75em}}Sylow Theorems}

\begin{enumerate}[label=\textbf{\sffamily\color{main}\Alph*.}]
	\item If $|G|=175$, prove that $G$ is abelian.
	\item Let $P$ be a Sylow $p$-subgroup of $G$. Prove $P$ is the only Sylow $p$-subgroup contained in the normalizer of $P$,
	$$N(P)=\{x\mid xPx^{-1}=P\}.$$
	\item Suppose $|G|=p^rm$, where $p$ is prime and $m$ is some integer with $p\not|m$, and suppose $H$ is a normal subgroup of $G$ with $|H|=p^\ell$. Prove $H$ is a subgroup of all Sylow $p$-subgroups of $G$.
\end{enumerate}